---
title: "Overview of the pgxsim package"
author: "Phil Chapman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pgxsimOverview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Dose response curves are ubiquitous in the biomedical sciences and drug discovery, and conventionally are used in preclinical drug discovery to compare the response of a biological system to a different compounds.  Such assays have been extensively validated and optimised to produce highly accurate point estimates of IC50 which can then be correlated with chemical differences between the compounds.

In oncology drug discovery, efforts have also been made to profile compounds in diverse biological models of interest.  Here the interest is not so much on explaining the difference in response between compounds, but the difference in response between models with different genetic characteristics.  An example of this approach is the profiling of over 200 compounds in ~1000 cancer cell lines the Genomics of Drug Sensitivity project carried out at the Wellcome Trust Sanger Centre.  

The challenge in this setting is that achieving assay consistency between different cell lines is difficult, and variation in the data results in loss of signal in terms of the effect that genetic covariates have on response to compound.  Reproducibility between different projects has been discussed here, here and here (REFS).

The pgxsim package provides functions to simulate dose response data given a known genetic effect on drug response.  It also provides functions to explore different methodologies to retrieve the genetic association.  Together these provide the basis to carry out simulation studies to inform experimental and analysis design.

# Set Up
## Load libraries
Load required libraries etc:
```{r, warning=FALSE, message=FALSE}
library(pgxsim)
library(tidyverse)
library(drc)
set.seed(10001)
```

## About the tidyverse
This vignette makes extensive use of the [tidyverse](http://www.tidyverse.org) suite of packages and concepts.  For an introduction see this [video on YouTube](https://www.youtube.com/watch?v=rz3_FDVt9eg) or read the [Managing Many Models section](http://r4ds.had.co.nz/many-models.html) of Hadley Wickham's R For Data Science book.. 

# Simulating data
## Simulation of cell lines with mutation data

First simulate 50 cell lines with a mutation at 20% frequency using the `sim_cell_lines` function:

```{r}
cl_data <- sim_cell_lines(n=100, type='discrete', prop=0.2)
head(cl_data)
cl_data %>% group_by(gene) %>% summarise(count=n())
```

Each cell line is scored 0 or 1 depending on whether it has the mutation or not.  Don't worry about the `cl_sd...` columns, these provide the ability to make dose response data from some cell lines more noisy than others.

## Simulation of pIC50 data

Having simulated some cell lines, we now want to simulate some estimates of pIC50.  This can be done using the `sim_pIC50` function as follows:

```{r}
sim_pIC50(mu=1, sd=0.1, beta=1, g=0, n=10, type='d')
```
Here we are providing an average value `mu`, a standard deviation `sd`, a `beta` value of 1 which is the size of the effect that we will eventually want to retrieve, and a value for `g` which determines whether or not the mutation is present.

To see how this works compare the following:
```{r}
mean(sim_pIC50(mu=1, sd=0.1, beta=1, g=0, n=10, type='d'))
mean(sim_pIC50(mu=1, sd=0.1, beta=1, g=1, n=10, type='d'))
```
The difference between the first (wild type) and second (mutant) set of cell lines is approximately 1 as expected from the value of `beta`.

To include the values in the cell line data frame we can use the function as follows:

```{r}
pIC50_data <- cl_data %>% 
  dplyr::mutate(pIC50 = sim_pIC50(mu=1, sd=0.1, beta=1, g=gene, n=nrow(cl_data), type='d'))
pIC50_data
```

We can check the difference between cell lines:
```{r}
pIC50_data %>% group_by(gene) %>% summarise(mean(pIC50))
ggplot(pIC50_data, aes(y=pIC50, x=as.factor(gene))) +
  geom_boxplot(aes(colour=as.factor(gene))) + theme_bw()
lm(pIC50 ~ as.factor(gene), data=pIC50_data)
```

To summarise, thus far we have simulated 100 cell lines with a mutation that exists at 20% frequency.  We have simulated pIC50 values for these cell lines given that the mutation confers a (average) 10 fold difference on sensitivity to drug (10 fold is 1 on a log10 scale).  

## Simulation of dose response data

We next need to simulate the experimental data itself, namely the response seen when the cell line is treated with compound at different concentrations.  The `sim_dose_response` function can be used to do this as follows:

```{r}
ex_dr1 <- sim_dose_response(pIC50=1, lb=0.001, ub=30, 
                            ndoses=10, nreps=1, sd_prop=0, sd_add=0)
ex_dr1
```
Here we specify the pIC50 value, the lowest concentration tested `lb`, the highest concentration tested `ub`, the number of doses `ndoses`, the number of replicates per dose `nreps`, and the additive and proportional variance `sd_prop` and `sd_add`.  Since we specified zero variance, we get the perfect dose response curve:

```{r}
ggplot(ex_dr1, aes(x=conc, y=resp)) + geom_point() + geom_line() + scale_x_log10() + theme_bw()
```

By changing `sd_add` and `sd_prop` we can introduce experimental noise into the data:

```{r}
ex_dr2 <- sim_dose_response(pIC50=1, lb=0.001, ub=30, 
                            ndoses=10, nreps=3, sd_prop=0.1, sd_add=0.1)
ggplot(ex_dr2, aes(x=conc, y=resp)) + geom_smooth(se = FALSE, method='loess') + 
  geom_point() + scale_x_log10() + theme_bw()

```

## Simulating data for multiple cell lines

Data for the first 10 cell lines we simulated earlier can be generated using the `purrr::map` function to generate a nested data frame as follows:

```{r}
dr_data <- pIC50_data %>%
  dplyr::filter(cell_id <= 10) %>%
  dplyr::mutate(data=purrr::map(pIC50, ~sim_dose_response(pIC50=., lb=0.001, ub=30, 
                            ndoses=10, nreps=3, sd_prop=0.1, sd_add=0.1)))
dr_data
dr_data$data[[1]]
```

This data frame can be unnested and the data plotted:

```{r}
dr_data_unnested <- tidyr::unnest(dr_data)
dr_data_unnested
ggplot(dr_data_unnested, aes(x=conc, y=resp)) + 
  geom_point() +
  scale_x_log10() +
  facet_wrap(~cell_id) +
  theme_bw()
```

## Simulating a full dataset

The functions that `pgxsim` provides are controlled by a variety of parameters, and we would like to explore the effect that these parameters have when trying to retrieve the genetic association.  To start with we want to specify a data frame containing all of the combinations of parameters that we wish to explore.  The `tidy::crossing` function is useful to achieve this, below we set up four simulations with different values for the proportional and additive variance, but the same principle can be applied to vary other parameters.

```{r}
param_df <- tidyr::crossing(lb=0.001, ub=30, ndoses=10, 
                            nreps=3, sd_prop=c(0.001, 0.2), 
                            sd_add=c(0.001, 0.2)) %>%
    dplyr::mutate(sim_no=row_number())
param_df
```

Next this data frame is crossed with the pIC50 data frame from earlier to give four simulations for each cell line:

```{r}
simulation_df <- tidyr::crossing(pIC50_data, param_df)
simulation_df
```

Finally we use the data in the data frame as parameters for the `sim_dose_response function:

```{r}
simulation_df <-  simulation_df %>% dplyr::mutate(
  dr_data=purrr::pmap(.l=list(pIC50, lb, ub, ndoses, nreps, sd_prop, sd_add),
                      .f=sim_dose_response)) %>%
  dplyr::select(sim_no, cell_id, gene, pIC50, sd_prop, sd_add, dr_data)
simulation_df %>% dplyr::select(sim_no, cell_id, dr_data)
```
The data for the first cell line is plotted below:

```{r}
simulation_df %>%
  dplyr::filter(cell_id==1) %>%
  tidyr::unnest() %>%
  ggplot(aes(x=conc, y=resp)) + geom_point() + scale_x_log10() + facet_wrap(~sim_no)
```

# Estimating pIC50 values from simulated data

## Using drc to estimate IC50's one-at-a-time

The `drc` package contains useful functions to fit conventional Hill curves to dose response data.  For example:

```{r}
fit <- drc::drm(resp~conc, fct=LL.4(), data=ex_dr2) 
plot(fit)
log(ED(fit, 50, display = FALSE))
```

This can be applied across multiple dose response datasets using `purrr::map` again:

```{r warning=FALSE, message=FALSE}
drc_results <- simulation_df %>%
  dplyr::mutate(fit=purrr::map(dr_data, ~drc::drm(resp~conc, fct=LL.4(), data=.)),
                drc_pIC50=purrr::map_dbl(fit, ~log(ED(., 50, display=FALSE)[1,'Estimate'])))
drc_results %>% dplyr::select(-dr_data, -fit)
```

Plotting the derived pIC50s against the original pIC50s shows how much the different amounts of experimental variability affects the accuracy of the curve fit:
```{r}
ggplot(drc_results, aes(x=pIC50, y=drc_pIC50))+
  geom_point() +
  theme_bw() +
  facet_wrap(~sim_no) +
  geom_abline(slope=1, intercept=1, colour='blue')
```

## Using mixed effects models to estimate IC50's simultaneously

Rather than estimating the IC50 values one by one, it is possible to estimate them in a single model fit using a non-linear mixed effects model (REF nlme).  For more information on this approach see a paper by Vis et al 2016 (REF).  Essentially the benefit here is that information on variability can be shared across cell lines which helps improve the accuracy of the fit.

Firstly, the nesting needs to be adjusted so that there is one row per simulation rather than one row per cell line.  This means that the data frame in the data list column represents data for all cell lines in that simulation which is what the `nlme_fit` function requires:

```{r}
nlme_data <- simulation_df %>%
  tidyr::unnest() %>%
  dplyr::group_by(sim_no, sd_prop, sd_add) %>%
  tidyr::nest()

nlme_data
```

The fit can then be done using the convenience function `nlme_fit`:

```{r}
nlme_res <- nlme_data %>%
  dplyr::mutate(fit=purrr::map(data, nlme_fit))
nlme_res
```
The pIC50 values obtained can be extracted using the convenience function `nlme_extract` and merged with the original pIC50 values:

```{r}
nlme_pIC50s <- nlme_res %>%
  dplyr::mutate(res=purrr::map(fit, nlme_extract)) %>%
  dplyr::select(-data,-fit) %>%
  tidyr::unnest() %>%
  dplyr::inner_join(pIC50_data, by='cell_id') %>%
  dplyr::select(-cl_sd_prop, -cl_sd_add)
nlme_pIC50s
```

Finally the estimated vs actual values can be plotted:

```{r}
ggplot(nlme_pIC50s, aes(x=pIC50, y=nlme_pIC50)) +
  geom_point() +
  theme_bw() +
  facet_wrap(~sim_no) +
  geom_abline(slope=1, intercept=1, colour='blue')
```


# Estimating genetic effects

## Using estimated IC50s in a linear regression

### Original data

Earlier in the simulation we defined the genetic covariate beta to equal 1, so a linear regression comparing mutated and wild type gives an estimate of around 1:

```{r}
lm(pIC50 ~ gene, data=pIC50_data)
ggplot(pIC50_data, aes(x=as.factor(gene), y=pIC50)) +
  geom_boxplot() +
  theme_bw()
```

### `drc` pIC50 estimates

To calculate the value of the genetic covariate from the pIC50 values estimated by `drc`, we can nest the `drc_results` data frame by simulation and carry out the linear regression on each simulation.  We then use `broom::tidy` to extract the genetic covariate estimate from the linear fit:

```{r}
drc_lm <- drc_results %>%
  dplyr::select(-fit, -dr_data) %>%
  dplyr::group_by(sim_no, sd_prop, sd_add) %>%
  tidyr::nest() %>%
  dplyr::mutate(lm_fit=purrr::map(data, ~lm(drc_pIC50 ~ gene, .)),
                lm_res=purrr::map(lm_fit, broom::tidy)) %>%
  dplyr::select(-lm_fit, -data) %>%
  tidyr::unnest() %>%
  dplyr::filter(term=='gene')
drc_lm
```

We see that our estimate of the genetic covariate is imprecise in the simulations where there is more noise in the data.

### `nlme` pIC50 estimates

We can carry out a similar analysis to that in the previous section on the nlme results:

```{r}
nlme_lm <- nlme_pIC50s %>%
  dplyr::group_by(sim_no, sd_prop, sd_add) %>%
  tidyr::nest() %>%
  dplyr::mutate(lm_fit=purrr::map(data, ~lm(nlme_pIC50 ~ gene, .)),
                lm_res=purrr::map(lm_fit, broom::tidy)) %>%
  dplyr::select(-lm_fit, -data) %>%
  tidyr::unnest() %>%
  dplyr::filter(term=='gene')
nlme_lm
```

Although the standard errors are smaller and the statistical test for the covariate being non-zero is significant, the estimate of the covariate value is still a long way from 1.  Multiple simulations with the same amount of experimental variations would have to be carried out to get an estimate of how accurate the estimates are.

## Including the genetic effect in the mixed effects model

In the previous section, recovering the genetic effect was a two stage process of first estimating pIC50 and then using this in a linear regression.  However, this approach discards the variability associated with the pIC50 point estimates.  This uncertainty can be retained by instead including the genetic covariate in the mixed effects model itself.  

The function `nlme_gene_fit` is used to fit this model:

```{r}
nlme_gene_res <- nlme_data %>%
  dplyr::mutate(fit=purrr::map(data, nlme_gene_fit))
nlme_gene_res
```

Since the genetic covariate is included in the model, the estimate can be retrieved directly:

```{r}
nlme_gene_res %>%
    dplyr::mutate(res=map(fit, broom::tidy, effects='fixed')) %>%
    dplyr::select(-data, -fit) %>%
    tidyr::unnest() %>%
    dplyr::filter(grepl('b', term)) 
```

Using this approach, the estimate of the genetic effect is near 1 for all simulations, even where a lot of experimental variability is present.

# Session Info
```{r}
sessionInfo()
```


